---
title: "Untitled"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
date: "2025-05-07"
---
```{r}
library(dplyr)
library(tidyr)    
library(knitr)
library(kableExtra)
library(ggplot2)
library(patchwork)
library(mgcv)
library(randomForest)
library(splines)
library(caret)
```

Data Preprocessing - Remove/Impute missing values, Remove outliers, Create 3-category income group
```{r}
freedom_byincome <- read.csv("/Users/audrey/Downloads/merged_data_income_class.csv")
summary(freedom_byincome)
```

```{r}
country_income_mode <- freedom_byincome %>%
  group_by(Country, Income_Group) %>%
  tally() %>%
  group_by(Country) %>%
  slice_max(order_by = n, n = 1, with_ties = FALSE) %>%
  dplyr::select(Country, income_group_mode = Income_Group)

freedom_income <- freedom_byincome %>%
  left_join(country_income_mode, by = "Country")

freedom_income <- freedom_income %>%
  mutate(
    income_group_3cat = case_when(
    income_group_mode == "High-income countries" ~ "high",
    income_group_mode %in% c("Upper-middle-income countries", 
                             "Lower-middle-income countries") ~ "middle",
    income_group_mode == "Low-income countries" ~ "low"
    )
  )
```

```{r}
missing_percent <- freedom_income %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100))

missing_percent <- pivot_longer(missing_percent, cols = everything(),
                                names_to = "variable", values_to = "missing_percentage")

print(missing_percent)
```

```{r}
# missing value > 40%
na_percent <- colMeans(is.na(freedom_income)) * 100
high_missing_vars <- names(na_percent[na_percent > 40])
high_missing_vars
```

```{r}
med_missing_vars <- names(na_percent[na_percent > 0])
med_missing_vars
```

```{r}
freedom_cleaned <- freedom_income %>%
  dplyr::select(-c(literacy_rate, Gini_index)) %>%
  filter(!is.na(HF)) %>%
  filter(
    cpi <= 100,
    Intentional_Homicides <= 100,
    infant_mortality <= 100
  )
```

```{r}
med_missing_vars <- c("SG", "Gender_Disparity_Index", "FTI", "gdp_per_capita", 
                      "cpi", "infant_mortality", "unemployment_rate", "intnet_use",
                      "Education_Expenditure", "Medical_Expenditure",
                      "Intentional_Homicides", "Political_Stability")

freedom_cleaned <- freedom_cleaned %>%
  mutate(across(all_of(med_missing_vars), ~ ifelse(is.na(.), median(., na.rm = TRUE), .)))
```


Linear Model
```{r}
freedom_cleaned$log_gdp <- log(freedom_cleaned$gdp_per_capita)
freedom_cleaned$income_group_3cat <- factor(freedom_cleaned$income_group_3cat)
freedom_cleaned$income_group_3cat <- relevel(freedom_cleaned$income_group_3cat, ref = "low")
group_counts <- table(freedom_cleaned$income_group_3cat)
freedom_cleaned$group_weight <- 1 / group_counts[freedom_cleaned$income_group_3cat]
freedom_cleaned$group_weight <- freedom_cleaned$group_weight / mean(freedom_cleaned$group_weight)
```

need interaction
```{r}
model_nointeract <- lm(HF ~ log_gdp + cpi + infant_mortality + unemployment_rate + 
                          intnet_use + Education_Expenditure + Medical_Expenditure +
                          Intentional_Homicides + Political_Stability + income_group_3cat,
                     data=freedom_cleaned)
```

```{r}
model_interact <- lm(HF ~ (log_gdp + cpi + infant_mortality + unemployment_rate + 
                          intnet_use + Education_Expenditure + Medical_Expenditure +
                          Intentional_Homicides + Political_Stability) * income_group_3cat,
                     data=freedom_cleaned, weights = group_weight)
anova(model_nointeract, model_interact)
summary(model_interact)
```


heteroskedasticity & non-linearity
```{r}
par(mfrow = c(2, 2))
plot(model_interact)
```

heteroskedasticity
```{r}
library(lmtest)
bptest(model_interact)
```

no severe multicollinearity
```{r}
library(car)
vif_values <- vif(model_interact, type = "predictor")
vif_df <- data.frame(vif_values)
vif_df
```

nonparametric extension

# LOESS to detect the nonlinearity
```{r,warning=FALSE}
vars <- c("log_gdp", "cpi", "infant_mortality", "unemployment_rate",
          "intnet_use", "Education_Expenditure", "Medical_Expenditure",
          "Intentional_Homicides", "Political_Stability")

plot_list <- lapply(vars, function(var) {
  ggplot(freedom_cleaned, aes(x = .data[[var]], y = log(HF))) +
    geom_point(alpha = 0.4, color = "black") +
    geom_smooth(method = "loess", se = TRUE, color = "blue") +
    labs(title = paste("HF vs", var), x = var, y = "Human Freedom Score") +
    theme_minimal()
})

wrap_plots(plotlist = plot_list, ncol = 3)
```

# Spline
```{r,warning=FALSE}
model_spline <- lm(HF ~ 
                     ns(log_gdp, df = 4) +
                     cpi +
                     ns(infant_mortality, df = 4) +
                     ns(unemployment_rate, df = 4) +
                     intnet_use +
                     ns(Education_Expenditure, df = 4) +
                     ns(Medical_Expenditure, df = 4) +
                     ns(Intentional_Homicides, df = 4) +
                     Political_Stability +
                     income_group_3cat,
                   data = freedom_cleaned, weights = group_weight)
```

# Spline with interact
```{r,warning=FALSE}
model_spline_interact <- lm(HF ~ 
  ns(log_gdp, df = 4) + 
  cpi + 
  ns(infant_mortality, df = 4) + 
  ns(unemployment_rate, df = 4) + 
  intnet_use + 
  ns(Education_Expenditure, df = 4) + 
  ns(Medical_Expenditure, df = 4) + 
  ns(Intentional_Homicides, df = 4) + 
  Political_Stability + 
  income_group_3cat +

  ns(infant_mortality, df = 4):income_group_3cat + 
  ns(unemployment_rate, df = 4):income_group_3cat + 
  ns(Education_Expenditure, df = 4):income_group_3cat + 
  ns(Medical_Expenditure, df = 4):income_group_3cat + 
  ns(Intentional_Homicides, df = 4):income_group_3cat +

  cpi:income_group_3cat +
  intnet_use:income_group_3cat +
  Political_Stability:income_group_3cat,
  
  data = freedom_cleaned, weights = group_weight
)

AIC(model_spline,model_spline_interact)
```

\newpage
```{r,echo=FALSE}
library(knitr)
spline_table <- as.data.frame(summary(model_spline_interact)$coefficients)

colnames(spline_table) <- c("Estimate", "Std. Error", "t-value", "p-value")

spline_table$Term <- rownames(spline_table)
rownames(spline_table) <- NULL

spline_table$Signif. <- cut(
  spline_table$`p-value`,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
  labels = c("***", "**", "*", ".", " "),
  right = TRUE
)

spline_table <- spline_table[, c("Term", "Estimate", "Std. Error", "t-value", "p-value", "Signif.")]

kable(spline_table, digits = 3, caption = "Table 3. Parametric coefficients from the spline interaction model.")

```

# Generalized Additive Model
```{r,warning=FALSE}
model_gam <- gam(HF ~ s(log_gdp) + cpi + s(infant_mortality) + s(unemployment_rate) +
                      intnet_use + s(Education_Expenditure) + s(Medical_Expenditure) +
                      s(Intentional_Homicides) + Political_Stability + income_group_3cat,
                       data = freedom_cleaned,weights = group_weight)
summary(model_gam)
```

# Add interaction term
```{r,warning=FALSE}
model_gam_interact <- gam(HF ~ 
  s(log_gdp) +
  s(infant_mortality) + s(infant_mortality, by = income_group_3cat) +
  s(unemployment_rate) + s(unemployment_rate, by = income_group_3cat) +
  s(Education_Expenditure) + s(Education_Expenditure, by = income_group_3cat) +
  s(Medical_Expenditure) + s(Medical_Expenditure, by = income_group_3cat) +
  s(Intentional_Homicides) + s(Intentional_Homicides, by = income_group_3cat) +
  cpi + intnet_use + Political_Stability +
  income_group_3cat +
  cpi:income_group_3cat + intnet_use:income_group_3cat + Political_Stability:income_group_3cat,
  data = freedom_cleaned, weights = group_weight
)

AIC(model_gam,model_gam_interact)
```
\newpage
```{r,echo=FALSE}
aic_compare <- AIC(model_spline, model_spline_interact,model_gam,model_gam_interact)
kable(aic_compare, digits = 2, caption = "Table 4. AIC comparison between spline models.")
```

# Random Forest regression
```{r,warning=FALSE}
rf_model <- randomForest(HF ~ log_gdp + cpi + infant_mortality +
                           unemployment_rate + intnet_use +
                           Education_Expenditure + Medical_Expenditure +
                           Intentional_Homicides + Political_Stability +
                           income_group_3cat,
                         data = freedom_cleaned,
                         importance = TRUE, ntree = 500)

pred_rf <- predict(rf_model)

rss_rf <- sum((freedom_cleaned$HF - pred_rf)^2)
tss <- sum((freedom_cleaned$HF - mean(freedom_cleaned$HF))^2) 
rsq_rf <- 1 - rss_rf / tss
rsq_rf
varImpPlot(rf_model)
```

# Cross validation for Spline
```{r,warning=FALSE}
set.seed(5291)
folds <- sample(rep(1:10, length.out = nrow(freedom_cleaned)))
cv_results_spline <- data.frame(Fold = 1:10, RMSE = NA, Rsq = NA)

for (k in 1:10) {
  train_data <- freedom_cleaned[folds != k, ]
  test_data  <- freedom_cleaned[folds == k, ]
  
  model_spline_interact <- lm(HF ~ 
    ns(log_gdp, df = 4) + 
    cpi + 
    ns(infant_mortality, df = 4) + 
    ns(unemployment_rate, df = 4) + 
    intnet_use + 
    ns(Education_Expenditure, df = 4) + 
    ns(Medical_Expenditure, df = 4) + 
    ns(Intentional_Homicides, df = 4) + 
    Political_Stability + 
    income_group_3cat +

    ns(infant_mortality, df = 4):income_group_3cat + 
    ns(unemployment_rate, df = 4):income_group_3cat + 
    ns(Education_Expenditure, df = 4):income_group_3cat + 
    ns(Medical_Expenditure, df = 4):income_group_3cat + 
    ns(Intentional_Homicides, df = 4):income_group_3cat +

    cpi:income_group_3cat +
    intnet_use:income_group_3cat +
    Political_Stability:income_group_3cat,
  
    data = train_data, weights = group_weight
)
  
  preds <- predict(model_spline_interact, newdata = test_data)
  actuals <- test_data$HF
  
  rss <- sum((preds - actuals)^2)
  tss <- sum((actuals - mean(actuals))^2)
  
  cv_results_spline$RMSE[k] <- sqrt(mean((preds - actuals)^2))
  cv_results_spline$Rsq[k] <- 1 - rss / tss
}

```

# Cross validation for GAM
```{r,warning=FALSE}
set.seed(5291)
folds <- sample(rep(1:10, length.out = nrow(freedom_cleaned)))

cv_results <- data.frame(Fold = 1:10, RMSE = NA, Rsq = NA)

for (k in 1:10) {
  train_data <- freedom_cleaned[folds != k, ]
  test_data  <- freedom_cleaned[folds == k, ]
  
  model_gam_interact <- gam(HF ~ 
  s(log_gdp) +
  s(infant_mortality) + s(infant_mortality, by = income_group_3cat) +
  s(unemployment_rate) + s(unemployment_rate, by = income_group_3cat) +
  s(Education_Expenditure) + s(Education_Expenditure, by = income_group_3cat) +
  s(Medical_Expenditure) + s(Medical_Expenditure, by = income_group_3cat) +
  s(Intentional_Homicides) + s(Intentional_Homicides, by = income_group_3cat) +
  cpi + intnet_use + Political_Stability +
  income_group_3cat +
  cpi:income_group_3cat + intnet_use:income_group_3cat + Political_Stability:income_group_3cat,
  data = train_data, weights = group_weight
)
  
  pred <- predict(model_gam_interact, newdata = test_data)
  obs <- test_data$HF
  
  rmse <- sqrt(mean((obs - pred)^2))
  rsq  <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
  
  cv_results[k, "RMSE"] <- rmse
  cv_results[k, "Rsq"] <- rsq
}

```

# Cross validation for random forest
```{r,warning=FALSE}
set.seed(5291)
cv_results_rf <- data.frame(Fold = 1:10, RMSE = NA, Rsq = NA)

for (k in 1:10) {
  train_data <- freedom_cleaned[folds != k, ]
  test_data  <- freedom_cleaned[folds == k, ]
  
  model_k <- randomForest(HF ~ log_gdp + cpi + infant_mortality +
                            unemployment_rate + intnet_use +
                            Education_Expenditure + Medical_Expenditure +
                            Intentional_Homicides + Political_Stability +
                            income_group_3cat,
                          data = train_data,
                          ntree = 500)
  
  pred <- predict(model_k, newdata = test_data)
  obs <- test_data$HF
  
  rmse <- sqrt(mean((obs - pred)^2))
  rsq  <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)
  
  cv_results_rf[k, "RMSE"] <- rmse
  cv_results_rf[k, "Rsq"] <- rsq
}
```

# Cross Validation for Linear model
```{r}
set.seed(5291)

cv_results_lm <- data.frame(Fold = 1:10, RMSE = NA, Rsq = NA)
folds <- sample(rep(1:10, length.out = nrow(freedom_cleaned)))

for (k in 1:10) {
  train_data <- freedom_cleaned[folds != k, ]
  test_data  <- freedom_cleaned[folds == k, ]

  model_k <- lm(HF ~ log_gdp + cpi + infant_mortality +
                  unemployment_rate + intnet_use +
                  Education_Expenditure + Medical_Expenditure +
                  Intentional_Homicides + Political_Stability +
                  income_group_3cat +
                  cpi:income_group_3cat +
                  infant_mortality:income_group_3cat +
                  unemployment_rate:income_group_3cat +
                  intnet_use:income_group_3cat +
                  Education_Expenditure:income_group_3cat +
                  Medical_Expenditure:income_group_3cat +
                  Intentional_Homicides:income_group_3cat +
                  Political_Stability:income_group_3cat,
                data = train_data, weights=group_weight)

  pred <- predict(model_k, newdata = test_data)
  obs  <- test_data$HF

  rmse <- sqrt(mean((obs - pred)^2))
  rsq  <- 1 - sum((obs - pred)^2) / sum((obs - mean(obs))^2)

  cv_results_lm[k, "RMSE"] <- rmse
  cv_results_lm[k, "Rsq"]  <- rsq
}

cv_results_lm

```

# Cross Validation Matrix
\newpage
```{r,warning=FALSE}
mean_spline <- colMeans(cv_results_spline[ , c("RMSE", "Rsq")])
mean_gam    <- colMeans(cv_results[ , c("RMSE", "Rsq")])
mean_rf     <- colMeans(cv_results_rf[ , c("RMSE", "Rsq")])
mean_lm     <- colMeans(cv_results_lm[ , c("RMSE", "Rsq")])
cv_summary <- rbind(Linear = mean_lm,
                    Spline = mean_spline,
                    GAM = mean_gam,
                    RF = mean_rf)
cv_summary
library(knitr)

kable(cv_summary, digits = 3, caption = "Table 5. 10-fold cross-validation results across models (mean RMSE and R-squared).")

```



```{r,echo=FALSE}
param_table <- as.data.frame(summary(model_gam_interact)$p.table)

colnames(param_table) <- c("Estimate", "Std. Error", "t-value", "p-value")

param_table$Term <- rownames(param_table)
rownames(param_table) <- NULL

param_table$Signif. <- cut(
  param_table$`p-value`,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
  labels = c("***", "**", "*", ".", " "),
  right = TRUE
)

param_table <- param_table[, c("Term", "Estimate", "Std. Error", "t-value", "p-value", "Signif.")]

kable(param_table, digits = 3, caption = "Table 1. Parametric coefficients from the GAM model.")

```

```{r,echo=FALSE}
library(knitr)
smooth_table <- as.data.frame(summary(model_gam_interact)$s.table)

colnames(smooth_table) <- c("edf", "Ref.df", "F", "p-value")

smooth_table$Term <- rownames(smooth_table)
rownames(smooth_table) <- NULL

smooth_table <- smooth_table[, c("Term", "edf", "Ref.df", "F", "p-value")]

kable(smooth_table, digits = 3, caption = "Table 2. Approximate significance of smooth terms in the GAM model.")

smooth_table$Signif. <- cut(
  smooth_table$`p-value`,
  breaks = c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
  labels = c("***", "**", "*", ".", " "),
  right = TRUE
)
kable(smooth_table, digits = 3, caption = "Table 2. Approximate significance of smooth terms in the GAM model.")


```


# Part 4: Inferential & Predictive Testing Procedures (Linear Regression)

library(car)
library(boot)
library(caret)

# 4.1 interactivity Model Comparison (ANOVA)
aov_comp <- anova(model_nointeract, model_interact, test = "Chisq")
print(aov_comp)

# 4.2 Coefficient Hypothesis Tests
predictors <- c(
  "log_gdp", "cpi", "infant_mortality", "unemployment_rate",
  "intnet_use", "Education_Expenditure", "Medical_Expenditure",
  "Intentional_Homicides", "Political_Stability"
)

# Per-predictor joint tests of middle & high interaction coefficients
for (p in predictors) {
  cat("\nTesting interactions for", p, ":\n")
  terms <- c(
    paste0(p, ":income_group_3catmiddle = 0"),
    paste0(p, ":income_group_3cathigh   = 0")
  )
  print(linearHypothesis(model_interact, terms, test = "F"))
}

# Joint test of all interaction terms simultaneously
all_ints <- grep(":", names(coef(model_interact)), value = TRUE)
hyp_all  <- paste0(all_ints, " = 0")
cat("\nJoint test of ALL interaction terms:\n")
print(linearHypothesis(model_interact, hyp_all, test = "F"))

# 4.3 Bootstrap Confidence Intervals for Key Coefficient
bootstrap_fn <- function(data, indices) {
  d <- data[indices, ]
  coef(
    lm(HF ~ log_gdp, data = d[d$income_group_3cat == 'low', ])
  )["log_gdp"]
}
set.seed(5291)
boot_res <- boot(freedom_cleaned, bootstrap_fn, R = 1000)
ci <- boot.ci(boot_res, type = "perc")
print(ci)

# 4.4 Predictive Calibration Plot for final_linear
library(dplyr)
library(ggplot2)

# get predicted values
freedom_cal <- freedom_cleaned %>%
  mutate(
    Pred_HF = predict(final_linear, newdata = freedom_cleaned),
    Bin = ntile(Pred_HF, 10)  # divide into 10 regions
  ) %>%
  group_by(Bin) %>%
  summarise(
    Mean_Pred = mean(Pred_HF),
    Mean_Obs  = mean(HF)
  )

ggplot(freedom_cal, aes(x = Mean_Pred, y = Mean_Obs)) +
  geom_point(size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    title = "Calibration Plot (Regression)",
    x = "Mean Predicted HF (by decile)",
    y = "Mean Observed HF (by decile)"
  ) +
  theme_minimal()
#*Interpretation:* compares predicted vs. observed quantiles of HF to assess systematic bias in predictions.
#-----------------------------------------------------------------------------------------------------
# Part4 Extension: Inferential & Predictive Tests (Nonparametric Models)

# --- 4.1 Spline Model: Test whether interacting splines improves fit ---
# Compare nested spline models via likelihood ratio (Chi-square) test
test_spline <- anova(model_spline, model_spline_interact, test = "Chisq")
print(test_spline)

# --- 4.2 GAM: Test significance of group-specific smooths ---
#  compare the base GAM vs. the interaction GAM
test_gam <- anova.gam(model_gam, model_gam_interact, test = "F")
print(test_gam)

# inspect approximate p-values of each smooth term
cat("\nSmooth term significance (model_gam_interact):\n")
print(summary(model_gam_interact)$s.table)

# --- 4.3 Random Forest: Permutation test for variable importance ---
# Permutation-based significance of RF variable importance
set.seed(5291)
rf_perm <- rfPermute(
  HF ~ log_gdp + cpi + infant_mortality + unemployment_rate + intnet_use +
    Education_Expenditure + Medical_Expenditure + Intentional_Homicides +
    Political_Stability + income_group_3cat,
  data = freedom_cleaned,
  ntree = 500,
  nrep = 100    
)
print(rf_perm)
# View p-values for each variable's importance
eval_perm <- importance(rf_perm)
print(eval_perm)

# --- 4.4 More Nonparametric Tests (Kruskal–Wallis & Dunn's post-hoc tests)  ---

library(dunn.test)     # Dunn’s post-hoc test
df <- freedom_cleaned

# Nonparametric group tests
# Boxplot for HF by income group
ggplot(df, aes(income_group_3cat, HF)) +
  geom_boxplot(fill="lightgray") +
  labs(title="HF by Income Group",
       x="Income Group", y="Human Freedom Score") +
  theme_minimal()

# Kruskal–Wallis & Dunn's post-hoc on HF
kruskal_hf <- kruskal.test(HF ~ income_group_3cat, data = df)
print(kruskal_hf)
dunn_hf <- dunn.test(df$HF, df$income_group_3cat,
                     method="bonferroni", alpha=0.05)
print(dunn_hf)


